Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-01-08 doc/invt.tex
editing invt.tex, add abstract

\documentclass[reprint]{revtex4-1}
\usepackage{amsmath}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{hyperref}

@@

@@

Given a system,
consider the problem of computing
the distribution, $p^*_i$
along a discrete quantity $i$.
%
The discrete nature of $i$ is natural
in discrete models.
%
For example, $i$ can be the energy $E$
in lattice models; or the temperature index
in a simulated tempering simulation.
%
For a continuous quantity, $z$,
which is typical in molecular systems,
we can also discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin, $(z, z + dz)$.
%
In both cases,
the distribution can be normalized as
$\sum_{i = 1}^n p^*_i = 1$.

For a large system,
the distribution $p^*_i$ is often
localized around some $i_0$,
%
and to find out the global property,
it is often desirable to carry out
a biased sampling that targets
a wider distribution $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

@@

Generally, we may wish to target
a slowly-varying distribution, $p_i$,
by adjusting the bias potential $\phi_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$\phi_i$ according to
$$
\phi_i^{\mathrm{new}}
=
\phi_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=\phi_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $\phi^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process can be inconvenient
in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

@@

Next, we split $h_i(t)$ into a deterministic part
and a noise
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
Here, the deterministic part can be related
to the ``ensemble average'' of $h_i$.
%
The ensemble consists of many similar simulation copies
that have experienced the same schedule $\alpha(t)$
and have reached the same $v_i(t)$
at time $t$.
%
The initial states and random numbers of the copies
may, however, be different.

@@

For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$.
%
%The noise is not necessarily white.
Unless specified otherwise, we assume
a general noise with the correlation
function depending on the time difference:
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.

@@

Variating this expression and
using the Euler-Lagrange equation yields
$$
\begin{aligned}
0
&=
\frac{d}{d\tau} \int_0^t
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt'
= \int_0^t
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by using Eq. \eqref{eq:kappat_limit}.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $\tau$,
we must have
%
\begin{equation}
\ddot u\bigl( q(t') \bigr) = 0,
\qquad
\mathrm{or}
\;\;
\dot u\bigl( q(t') \bigr) = c,
\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:ut_definition}
we get
$$
e^{-\lambda \, q(t) + \lambda \, q(t')}
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm, and differentiating this with respect to $t'$
yields Eq. \eqref{eq:alpha_opt}.

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin updating scheme,
we find from Eq. \eqref{eq:dxdt_WL}
that with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
Thus, Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$ defining the origin of $t$.

@@

We now extend the above analysis to
the multiple-bin updating scheme.
%
In this case,
upon a visit to bin $j$,
we update not only $v_j$,
but also $v_i$ for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

The procedure of optimization is largely
the same as the single-bin case,
whereas the matrix nature brings about
some difficulties that require
certain restrictions and approximations.

@@

\subsection{Comparison of updating schemes}

A more direct, and often more accurate, approach
is to assume the functional form
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
and find the optimal value of $\lambda$.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_mbin_perfect} yield
%
\begin{equation}
\mathcal E
\stackrel{t \to \infty}
{=\joinrel=\joinrel=}
\frac{1}{t}
\sum_{i = 2}^n
\frac{ \lambda_i^2 } { \lambda \, (2 \, \lambda_i - \lambda) },
\label{eq:y2_sum_perfect_invt}
\end{equation}
%
and the optimal $\lambda$
can be found from the smallest real root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

@@

This matrix is clearly symmetric, and hence
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution $p_i \equiv 1/n$.
%
It admits $n$ eigenvectors,
$\pmb\theta^{(1)}, \dots, \pmb\theta^{(n)}$,
given by
%
\begin{equation}
\theta^{(k)}_i
= \cos \frac{ \left(i - \frac12\right) (k - 1) \, \pi}{n}.
\label{eq:wband_eigenvector}
\end{equation}
%
with eigenvalues
\begin{equation}
  \lambda_k = m_0 + \sum_{l = 1}^b 2 \, m_l \cos \frac{(k - 1)  \, l \pi}{n}.
\label{eq:wband_eigenvalue}
\end{equation}

@@

Now
$$
\begin{aligned}
  &
  \sum_{i=1}^n
  \theta^{(k)}_i \,
  \left(
    m_{i - j} + m_{2n+1-i-j}
  \right)
  \\
  &=
  \sum_{l=1-j}^{n-j} \theta^{(k)}_{j+l} \, m_l
  +
  \sum_{l=n-j+1}^{2n-j} \theta^{(k)}_{2n+1-j-l} \, m_l
  \\
  &=
  \sum_{l=1-j}^{2n-j} \theta^{(k)}_{j+l} \, m_l
  =
  \sum_{l=0}^{b} m_l \, \cos\frac{(j+l-\frac12)(k-1)\pi}{n},
\end{aligned}
$$
where we have used
$$
\theta^{(k)}_i = \theta^{(k)}_{1-i} = \theta^{(k)}_{2n+1-i}.
$$
Similarly,
$$
\begin{aligned}
  \sum_{i=1}^n
  \theta^{(k)}_i
  \left(
    m_{j - i} + m_{i+j-1}
  \right)
  =
  \sum_{l=0}^{b} m_l \, \cos\frac{(j-l-\frac12)(k-1)\pi}{n}.
\end{aligned}
$$
%
So
$$
\begin{aligned}
\sum_{i = 1}^n \theta^{(k)}_i \, w_{ij}
&=
- \theta^{(k)}_j \, m_0
+\sum_{l=0}^{b} m_l \, 2 \, \cos\frac{(k-1)l\pi}{n} \, \cos\frac{(j-\frac12)(k-1)\pi}{n}
\\
&= \theta^{(k)}_j \, \lambda_k,
\end{aligned}
$$
with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue}.
}
%

@@

Particularly, if $m_2 = \dots = m_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%
\begin{equation}
\arraycolsep=1.4pt\def\arraystretch{1.4}
\mathbf w
=
\left(
  \begin{array}{cccccccc}
    1 - m_1   & m_1 & 0 & \dots & 0 \\
    m_1 & 1 - 2 \, m_1  & m_1 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & m_1 & 1 - 2 \, m_1  & m_1 \\
    0 & \dots & 0 & m_1 & 1 - m_1
  \end{array}
\right).
\label{eq:wtridiag}
\end{equation}
%
We have
\begin{equation}
  \lambda_k = 1 - 4 \, m_1 \sin^2 \frac{(k - 1) \, \pi}{2n}.
\label{eq:wtridiag_eigenvalue}
\end{equation}

Thus, this updating matrix requires
$$
m_1 < \frac{1}{4 \, \cos^2\frac{\pi}{2n} }.
$$

For a large $n \gg 1$,
we can convert Eq.
\eqref{eq:y2_sum_perfect_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$$
\begin{aligned}
\mathcal E
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, m_1 \, \cos^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, m_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, m_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{
  The integral is done by contour integration.
  %
$$
\begin{aligned}
\mathcal E
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, m_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - m_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, m_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is around the unit circle in the complex plane.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, m_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, m_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, m_1 - \lambda}}
{\sqrt{8 m_1}}.
$$
Thus,
$$
\begin{aligned}
  \mathcal E
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - m_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 m_1 - \lambda - 4 m_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ \left(1 - m_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 m_1 - \lambda - 4 m_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 m_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, m_1} { 1 - 2 \, m_1 },
\label{eq:lambda_tridiag}
\end{equation}
%
and the minimal error is given by
%
\begin{equation}
\mathcal E
=
\frac{n}{t}
\left(
  1+ \frac{2 \, m_1^2}{1-4 \, m_1}
\right),
\label{eq:error_tridiag}
\end{equation}
%
which increases rapidly with $m_1$.

@@

Another common updating scheme
in metadynamics is a Gaussian wave.
%
With $b = n - 1 \gg 1$,
we can take the continuous limit:
$x = l \, \Delta x$,
$\Delta x = \pi/n$,
and
$m_l = m(x) \, \Delta x$.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^{\pi}
m(x) \, \cos\bigl((k-1) \, x\bigr) \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^{\pi} m(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

Particularly, for the Gaussian wave
%
\begin{equation}
m(x) = \frac{1}{\sqrt{2\pi\sigma^2}}
\exp\left( -\frac{ x^2 } { 2 \, \sigma^2 } \right),
\end{equation}
%
with $\Delta x \ll \sigma \ll 1$,
we can extend the upper limit of the integrals
in Eqs. \eqref{eq:lambda_int}
and \eqref{eq:mx_normalization} to infinity,
and
\begin{equation}
\lambda_k = \exp\left(
  -\frac{ (k-1)^2 \, \sigma^2 } { 2 }
\right).
\label{eq:lambda_Gaussian}
\end{equation}

Owing to singularity,
we can show that the optimal $\lambda$
in Eq. \eqref{eq:alpha_trial}
\begin{equation}
  \lambda \approx 2 \lambda_{n}.
\end{equation}

>>>>

\documentclass[reprint]{revtex4-1}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{hyperref}

@@

\begin{abstract}
  Wang-Landau (WL) and metadynamics algorithms
  are widely-used techniques in free energy simulations.
  %
  These techniques allow one to sample a flat distribution
  along a quantity of interest
  by constructing a bias potential on the fly,
  and to compute the negative potential of mean force (PMF),
  which is the negative of the bias potential.
  %
  However, the asymptotic error of the resulting PMF
  is determined by the magnitude of updating the bias potential,
  and in what fashion the magnitude is reduced
  over the simulation time.
  %
  In the Wang-Landau case,
  an effective prescription of determining the magnitude
  is to set it to be the inverse of the number of simulation steps.
  %
  Here, we shall reexamine the optimality of this prescription,
  and attempt to extend it to the metadynamics case,
  in which the bias potential at several neighboring bins
  are updating according to a prescribed function,
  upon each visit.
  %
  Our analysis also suggests that in terms of the convergence
  the WL algorithm represents one of the best of the class
  updating schemes among more sophisticated candidates.
\end{abstract}

@@

Given a system,
consider the problem of computing
the distribution, $p^*_i$
along a discrete quantity $i$.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy $E$
in lattice spin models; or the temperature index
in a simulated tempering simulation.
%
For the quantity of interest, $z$, is continuous,
which is typical in molecular systems,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin, $(z, z + dz)$.
%
In both cases,
the distribution can be normalized as
$\sum_{i = 1}^n p^*_i = 1$.

For a large system,
the distribution $p^*_i$ tends to
be localized around some $i_0$,
%
and to find out the global property,
it is often desirable to carry out
a biased sampling that targets
a wider distribution $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

@@

Generally, we may wish to sample
a nearly flat distribution, $p_i$,
by adjusting the bias potential $\phi_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$\phi_i$ according to
$$
\phi_i^{\mathrm{new}}
=
\phi_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=\phi_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $\phi^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process can be inconvenient
in practice,
for we need to make the period
as long as a multiple of
the autocorrelation time,
which is difficult to estimate for an unknown system.

@@

Next, we split $h_i(t)$ into a deterministic part
and a noise part,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
Here, the deterministic part can be related
to the ``ensemble average'' of $h_i$.
%
The ensemble consists of many similar simulation copies
that have experienced the same schedule $\alpha(t)$
and have reached the same $v_i(t)$
at time $t$.
%
The initial states and random numbers of the copies
may, however, be different.

@@

For the noise part, we have
$\langle \zeta_i(t) \rangle = 0$.
%
%The noise is not necessarily white.
We consider a general noise with the correlation
function depending on the time difference:
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function of $t$
that vanishes at large $t$.

@@

Variating this expression and
using the Euler-Lagrange equation yields
$$
\begin{aligned}
0
&=
\frac{d}{d\tau} \int_0^t
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt'
\\
&= \int_0^t
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by using Eq. \eqref{eq:kappat_limit}.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $\tau$,
we must have
%
\begin{equation}
\ddot u\bigl( q(t') \bigr) = 0,
\qquad
\mathrm{or}
\;\;
\dot u\bigl( q(t') \bigr) = c,
\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:ut_definition}
we get
$$
e^{-\lambda \, q(t) + \lambda \, q(t')}
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm, and differentiating this with respect to $t'$
yields Eq. \eqref{eq:alpha_opt}.

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin updating scheme,
we find from Eq. \eqref{eq:dxdt_WL}
that with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
Thus, Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$. %defining the origin of $t$.

@@

We now consider the multiple-bin updating scheme.
%
In this case,
upon a visit to bin $j$,
we update not only the bias potential there,
but also those for a few neighboring bins $i$.
%
The update magnitudes are given by a matrix $\mathbf w$
with elements $w_{ij}$.
%
Then, Eq. \eqref{eq:vt_diffeq} is generalized to
\begin{equation}
  \dot v_i(t) =
  \sum_{j=1}^n \alpha(t) \, h_j(t) \frac{ w_{ij} } { p_j }.
  \label{eq:vt_diffeq_mbin}
\end{equation}

The procedure of optimization is largely
the same as the single-bin case,
whereas the matrix nature brings about
some technical difficulties that require
certain restrictions and approximations.

@@

\subsection{Optimal coefficient of proportionality}

A more direct, and often more accurate, approach
is to assume the functional form
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_trial}
\end{equation}
%
and to find the optimal value of $\lambda$.
%
Using Eq. \eqref{eq:alpha_trial}
in Eq. \eqref{eq:error_mbin_perfect} yields
%
\begin{align}
\mathcal E
&=
\frac{1}{t+t_0}
\sum_{i=2}^n \frac{ \lambda_i^2 } { \lambda \, (2 \lambda_i - \lambda) }
\left[
  1 - \left(
    1 + \frac{t}{t_0}
  \right)^{1 - 2 \, \lambda_i/\lambda}
\right]
\notag
\\
&\stackrel{      t \gg t_0      }
          { =\joinrel=\joinrel= }
%
\sum_{i = 2}^n
\begin{dcases}
  \frac{ 1 }
       { t }
  \frac{ \lambda_i^2 }
       { \lambda \, (2 \, \lambda_i - \lambda) }
  &
  \mathrm{if \;} \lambda < 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{   1    }
       { 4 \, t }
  \ln \frac{ t + t_0 }
           {  t_0    }
  &
  \mathrm{if \;} \lambda = 2 \, \lambda_i,
  \\%[1em]
  %
  %
  \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1} }
       { t^{ 2 \, \lambda_i / \lambda } }
  \frac{ \lambda_i^2 }
       { \lambda \, (\lambda - 2 \, \lambda_i) }
  &
  \mathrm{if \;} \lambda > 2 \, \lambda_i.
\end{dcases}
\label{eq:y2_sum_perfect_invt}
\end{align}
%
\note{In the above expression,
$t + t_0$ are replaced by $t$.
%
Adding back $t_0$, we get
$$
\begin{aligned}
  \mathcal E
  \stackrel{         t \to \infty         }
           { =\joinrel=\joinrel=\joinrel= }
  \sum_{i = 2}^n
  \begin{dcases}
    \frac{ 1 }{ t + t_0 }
    \frac{ \lambda_i^2 } { \lambda \, (2 \, \lambda_i - \lambda) }
    &
    \mathrm{if \;} \lambda < 2 \, \lambda_i,
    \\%[1em]
    %
    %
    \frac{1}{4 \, (t + t_0) } \ln\frac{t + t_0}{t_0}
    &
    \mathrm{if \;} \lambda = 2 \, \lambda_i,
    \\
    %
    %
    \frac{ t_0^{ 2 \, \lambda_i / \lambda  - 1} }
         { (t + t_0)^{ 2 \, \lambda_i / \lambda } }
    \frac{ \lambda_i^2 }
         { \lambda \, (\lambda - 2 \, \lambda_i) }
    &
    \mathrm{if \;} \lambda > 2 \, \lambda_i.
  \end{dcases}
\end{aligned}
$$
}
%
The last two cases, $\lambda \ge 2\lambda_i$,
are asymptotically suboptimal
as they decay slower than $1/t$.
%
Thus, the optimal $\lambda$
satisfies $\lambda < 2 \min \lambda_i$,
and in this case,
we can seek that
the optimal value from
the smallest real root of
%
\begin{equation}
\sum_{i = 2}^n
\frac{ \lambda_i - \lambda }
{ \left(2 - \lambda/ \lambda_i \right)^2 }
= 0.
\label{eq:optimal_lambda_approx}
\end{equation}

@@

This matrix, being symmetric,
satisfies Eqs. \eqref{eq:w_sumj}-\eqref{eq:w_balance},
for the flat distribution $p_i \equiv 1/n$.
%
It admits $n$ eigenvectors,
$\pmb\theta^{(1)}, \dots, \pmb\theta^{(n)}$,
given by
%
\begin{equation}
\theta^{(k)}_i
= \cos \frac{ \left(i - \frac12\right) (k - 1) \, \pi}{n}.
\label{eq:wband_eigenvector}
\end{equation}
%
with eigenvalues
\begin{equation}
  \lambda_k = m_0 + \sum_{l = 1}^b 2 \, m_l \cos \frac{(k - 1)  \, l \pi}{n}.
\label{eq:wband_eigenvalue}
\end{equation}

@@

Now
$$
\begin{aligned}
  &
  \sum_{i=1}^n
  \theta^{(k)}_i \,
  \left(
    m_{i - j} + m_{2n+1-i-j}
  \right)
  \\
  &=
  \sum_{l=1-j}^{n-j} \theta^{(k)}_{j+l} \, m_l
  +
  \sum_{l=n-j+1}^{2n-j} \theta^{(k)}_{2n+1-j-l} \, m_l
  \\
  &=
  \sum_{l=1-j}^{2n-j} \theta^{(k)}_{j+l} \, m_l
  =
  \sum_{l=0}^{b} m_l \, \cos\frac{(j+l-\frac12)(k-1)\pi}{n},
\end{aligned}
$$
where we have used
$$
\theta^{(k)}_i = \theta^{(k)}_{1-i} = \theta^{(k)}_{2n+1-i}.
$$
Similarly,
$$
\begin{aligned}
  \sum_{i=1}^n
  \theta^{(k)}_i
  \left(
    m_{j - i} + m_{i+j-1}
  \right)
  =
  \sum_{l=0}^{b} m_l \, \cos\frac{(j-l-\frac12)(k-1)\pi}{n}.
\end{aligned}
$$
%
So
$$
\begin{aligned}
\sum_{i = 1}^n \theta^{(k)}_i \, w_{ij}
&=
- \theta^{(k)}_j \, m_0
+\sum_{l=0}^{b} m_l \, 2 \, \cos\frac{(k-1)l\pi}{n} \, \cos\frac{(j-\frac12)(k-1)\pi}{n}
\\
&= \theta^{(k)}_j \, \lambda_k,
\end{aligned}
$$
with $\lambda_k$ given by Eq. \eqref{eq:wband_eigenvalue}.
}
%
We discuss two special cases below.

@@

Particularly, if $m_2 = \dots = m_b = 0$,
the matrix $\mathbf w$ is tridiagonal,
%
\begin{equation}
\arraycolsep=1.4pt\def\arraystretch{1.4}
\mathbf w
=
\left(
  \begin{array}{cccccccc}
    1 - m_1   & m_1 & 0 & \dots & 0 \\
    m_1 & 1 - 2 \, m_1  & m_1 & \dots & 0 \\
    \vdots & &  & & \vdots \\
    0 & \dots & m_1 & 1 - 2 \, m_1  & m_1 \\
    0 & \dots & 0 & m_1 & 1 - m_1
  \end{array}
\right).
\label{eq:wtridiag}
\end{equation}
%
We have
\begin{equation}
  \lambda_k = 1 - 4 \, m_1 \sin^2 \frac{(k - 1) \, \pi}{2n}.
\label{eq:wtridiag_eigenvalue}
\end{equation}
%
Thus, for this updating matrix to be stable,
we need
$\min \lambda_k = \lambda_n > 0$,
and
$$
m_1 < \frac{1}{4 \, \cos^2\frac{\pi}{2n} }.
$$

We can find an analytical expression
of the optimal value of $\lambda$
in the limit of large $n$.
%
In this case,
we convert Eq.
\eqref{eq:y2_sum_perfect_invt}
%and
%\eqref{eq:optimal_lambda_approx}
to an integral,
$$
\begin{aligned}
\mathcal E
&=
\frac{2 \, n}{\pi \, t}
\int_0^{\pi/2}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ \lambda \, \left(2 - 8 \, m_1 \, \cos^2 p - \lambda \right) }
\, dp
\notag \\
&=
\frac{n}{4 \, t}
\left(
  \frac{2 - 4 \, m_1 + \lambda}{ \lambda }
  +
  \frac{ \lambda }
  { \sqrt{ (2 - \lambda) (2 - 8 \, m_1 - \lambda) } }
\right).
\end{aligned}
$$
\note{The integral is evaluated by contour integration.
  %
$$
\begin{aligned}
\mathcal E
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\int_0^{2 \, \pi}
\frac{ \left(1 - 4 \, m_1 \, \cos^2 p \right)^2 }
{ 2 - 8 \, m_1 \, \cos^2 p - \lambda }
\, dp
\\
&=
\frac{n}{\lambda \, t}
\frac{1}{2 \pi i}
\oint
\frac{ \left[1 - m_1 \, \left(z+\frac{1}{z}\right)^2 \right]^2 }
{ 2 - 2 \, m_1 \, \left(z + \frac{1}{z}\right)^2 - \lambda }
\, \frac{dz}{z},
\end{aligned}
$$
where the contour is the unit circle
in the complex plane of $z$.
%
This integral has five poles, one at the origin,
which produces a residue $\frac{2 - 4 \, m_1 - \lambda}{4}$
(from series expansion of small $z$),
and the other four at
$$
z + \frac{1}{z} = \pm\sqrt\frac{2-\lambda}{2 \, m_1},
$$
with two of them inside the unit circle:
$$
z_\pm = \pm \frac{\sqrt{2-\lambda} -\sqrt{2 - 8 \, m_1 - \lambda}}
{\sqrt{8 m_1}}.
$$
Thus,
$$
\begin{aligned}
  \mathcal E
&=
\frac{n}{\lambda \, t}
\left(
 \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \sum_{z = z_{\pm} }
 \frac{ \left(z^2 - m_1 (1 + z^2)^2 \right)^2 }
 { 2 z^4 (2 - 4 m_1 - \lambda - 4 m_1 z^2) }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ \left(1 - m_1 \left(z + \frac{1}{z} \right)^2 \right)^2 }
 { 2 - 4 m_1 - \lambda - 4 m_1 z_{\pm}^2 }
\right)
\\
&=
\frac{n}{\lambda \, t}
\left(
  \frac{2 - 4 \, m_1 - \lambda}{4}
 +
 \frac{ (\lambda/2)^2 }
 { \sqrt{(2-\lambda) (2 - 8 m_1 -\lambda)} }
\right).
\end{aligned}
$$
}
%
Minimizing the function yields
%
\begin{equation}
\lambda = \frac{1 - 4 \, m_1} { 1 - 2 \, m_1 },
\label{eq:lambda_tridiag}
\end{equation}
%
and the error in the optimal is given by
%
\begin{equation}
\mathcal E
=
\frac{n}{t}
\left(
  1+ \frac{2 \, m_1^2}{1-4 \, m_1}
\right).
\label{eq:error_tridiag}
\end{equation}
%
From Eq. \eqref{eq:error_tridiag},
it is clear that the error
increases rapidly with magnitude of $m_1$,
no matter its sign,
and optimal value is given by $m_1 = 0$,
which is the single-bin update scheme.

@@

Another common updating scheme
in metadynamics is a Gaussian wave.
%
Below we show that the updating scheme
is stable, although
it might be less effective
in reducing local noises.

We start by defining
a continuous limit
for $b = n - 1 \gg 1$:
$x = l \, \Delta x$,
$\Delta x = \pi/n$,
and
$m_l = m(x) \, \Delta x$.
%
Then
Eq. \eqref{eq:wband_eigenvalue}
can be approximated by an integral:
%
\begin{equation}
\lambda_k
=
2 \int_0^\pi
  m(x) \, \cos \bigl( (k-1) \, x \bigr) \, dx,
\label{eq:lambda_int}
\end{equation}
%
with the normalization
%
\begin{equation}
1 = 2 \int_0^\pi m(x) \, dx.
\label{eq:mx_normalization}
\end{equation}

For the Gaussian wave,
$m(x) \propto \exp(-\frac{1}{2} x^2/\sigma^2)$.
%
If $\Delta x \ll \sigma \ll 1$,
we can extended
the upper limit of the integrals
in Eqs. \eqref{eq:lambda_int}
and \eqref{eq:mx_normalization}
to infinity, and
%
\begin{equation}
m(x)
=
\frac{            1            }
     { \sqrt{ 2 \pi \sigma^2 } }
%
\exp\left(
      -\frac{       x^2     }
            { 2 \, \sigma^2 }
    \right),
\end{equation}
%
%
The eigenvalues are given by
%
\begin{equation}
\lambda_k
=
\exp\left(
      -\frac{ (k - 1)^2 \, \sigma^2 }
            {           2           }
    \right).
\label{eq:lambda_Gaussian}
\end{equation}
%
The eigenvalues are positive,
which means that the updating scheme
is stable.
